Budgeted Reinforcement Learning in Continuous State Space

### 2019-10-26 paper/main.tex
Updates from Overleaf

\begin{document}
\maketitle
\begin{abstract}
    A Budgeted Markov Decision Process (BMDP) is an extension of a Markov Decision Process to critical applications requiring safety constraints. It relies on a notion of risk implemented in the shape of a cost signal constrained  to lie below an -- adjustable -- threshold.
    So far, BMDPs could only be solved in the case of finite state spaces with known dynamics. This work extends the state-of-the-art to continuous spaces environments and unknown dynamics. We show that the solution to a BMDP is a fixed point of a novel Budgeted Bellman Optimality operator. This observation allows us to introduce natural extensions of Deep Reinforcement Learning algorithms to address large-scale BMDPs. We validate our approach on two simulated applications: spoken dialogue and autonomous driving\footnote{Videos and code are available at \href{https://budgeted-rl.github.io/}{this https URL} as required by the \href{https://docs.google.com/document/d/1SBCfisdLU2A_vb7K7wI8oRBin8ZA4cA1BbasqfQr5m0/edit}{Code Submission Policy}.}.
\end{abstract}

@@

\section{Budgeted Dynamic Programming}
\label{sec:bdp}
 We work in the space of budgeted policies, where $\pi$ both depends on $\beta$ and also outputs a next budget $\beta_a$. Hence, the budget $\beta$ is neither fixed nor constant as in the CMDP setting but instead evolves as part of the dynamics.

@@

\begin{proposition}[Optimality of the greedy policy]
\label{prop:greedy_optimal}
The greedy policy $\pi_\text{greedy}(\cdot~; Q^*)$ is \emph{uniformly} optimal: for all $\os\in\ocS$, $\pi_\text{greedy}(\cdot~; Q^*)\in\Pi^*(\os)$. In particular, $V^{\pi_\text{greedy}(\cdot; Q^*)} = V^*$ and $Q^{\pi_\text{greedy}(\cdot; Q^*)}= Q^*$.
\end{proposition}

@@

\begin{theorem}[Contractivity of $\cT$ on smooth $Q$-functions]
\label{rmk:contractivity-smooth}
The operator $\cT$ is a contraction when restricted to the subset $\cL_\gamma$ of $Q$-functions such that "$Q_r$ is $L$-Lipschitz with respect to $Q_c$", with $L<\frac{1}{\gamma}-1$. %We lengthily discuss some intuition on why that should be the case in \Cref{proof:contraction-with-smooth}.
\end{theorem}

\begin{remark}
Because of the budget dynamics $\beta' = \beta_a$ in \eqref{eq:dynamics}, we can see in \eqref{eq:bellman_expectation} that for all $\os=(s, \beta)$ and $\oa=(a, \beta_a)$, $Q^\pi(\os, \oa)$ only depends on $s, a, \beta_a$ and not on $\beta$. We will slightly abuse notations and sometimes denote in the following $Q^\pi(s, a, \beta_a) \eqdef Q^\pi(\os, \oa)$.
\end{remark}

@@

We first bring to light the role of risk-sensitive exploration in the \text{corridors} environment: \Cref{fig:exploration} shows the set of trajectories collected by each exploration strategy\footnote{Animations are available at \href{https://budgeted-rl.github.io/\#risk-sensitive-exploration}{this https URL}}, 
%in \Cref{subsec:exploration-examples}}
and the resulting performance of a budgeted policy trained on each batch. The trajectories (orange) in the risk-neutral batch are concentrated along the risky corridor (red) and ignore the safe corridor (green), which results in bad performances in the low-risk regime. Conversely, trajectories in the risk-sensitive batch (blue) are well distributed among both corridors and the corresponding budgeted policy achieves good performance across the whole spectrum of risk budgets.

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.23\textwidth]{source/img/test.pdf}
    %\includegraphics[width=0.23\textwidth]{source/img/risk-sensitive.png}
    \includegraphics[page=1, width=0.45\textwidth]{source/img/corridors}
    \caption{Trajectories (left) and performances (right) of two exploration strategies in the \texttt{corridors} environment. }
    \label{fig:exploration}
\end{figure}

@@

In a second experiment displayed in \Cref{fig:results}, we compare the performance of \FTQl to that of \BFTQ in the dialogue and autonomous driving tasks. For each algorithm, we plot the reward-cost trade-off curve. In both cases, \BFTQ performs almost as well as \FTQl despite only requiring a single model. All budgets are well-respected on \text{slot-filling}, but on \text{highway-env} we can observe an underestimation of $Q_c$, since e.g. $\expectedvalue[G_c|\beta=0] \simeq 0.1 $. This underestimation can be a consequence of two approximations: the use of the sampling operator $\hat{\cT}$ instead of the true population operator $\cT$, and the use of the neural network function approximation $Q_\theta$ instead of $Q$. 
% This is probably due to \text{highway-env} being actually infeasible for $\beta=0$, since we noticed that about $8\%$ of bad initialisations occur with the front vehicle driving slower than the minimum velocity of the controlled vehicle, leading to either a front collision or a forced change to the opposite lane, both yielding costs.
Still, \BFTQ provides a better control on the expected cost of the policy, than \FTQl. In addition, \BFTQ behaves more consistently than \FTQl overall, as shown by its lower extra-seed variance. Examples of policy execution can be found at \href{https://budgeted-rl.github.io/\#optimal-budgeted-policies}{this https URL}.
%in \Cref{sec:bftq-executions}.

>>>>

\begin{document}
\maketitle
\begin{abstract}
    A Budgeted Markov Decision Process (BMDP) is an extension of a Markov Decision Process to critical applications requiring safety constraints. It relies on a notion of risk implemented in the shape of a cost signal constrained  to lie below an -- adjustable -- threshold.
    So far, BMDPs could only be solved in the case of finite state spaces with known dynamics. This work extends the state-of-the-art to continuous spaces environments and unknown dynamics. We show that the solution to a BMDP is a fixed point of a novel Budgeted Bellman Optimality operator. This observation allows us to introduce natural extensions of Deep Reinforcement Learning algorithms to address large-scale BMDPs. We validate our approach on two simulated applications: spoken dialogue and autonomous driving\footnote{Videos and code are available at \url{https://budgeted-rl.github.io/}%{this https URL}
    %as required by the \href{https://docs.google.com/document/d/1SBCfisdLU2A_vb7K7wI8oRBin8ZA4cA1BbasqfQr5m0/edit}{Code Submission Policy}
    .}.
\end{abstract}

@@

\section{Budgeted Dynamic Programming}
\label{sec:bdp}
 We work in the space of budgeted policies, where a policy $\pi$ both depends on the current budget $\beta$ and also outputs a next budget $\beta_a$. Hence, the budget $\beta$ is neither fixed nor constant as in the CMDP setting but instead evolves as part of the dynamics.

@@

\begin{proposition}[Optimality of the greedy policy]
\label{prop:greedy_optimal}
The greedy policy $\pi_\text{greedy}(\cdot~; Q^*)$ is \emph{uniformly} optimal: $\forall\os\in\ocS$, $\pi_\text{greedy}(\cdot~; Q^*)\in\Pi^*(\os)$. In particular, $V^{\pi_\text{greedy}(\cdot; Q^*)} = V^*$ and $Q^{\pi_\text{greedy}(\cdot; Q^*)}= Q^*$.
\end{proposition}

@@

\begin{theorem}[Contractivity of $\cT$ on smooth $Q$-functions]
\label{rmk:contractivity-smooth}
The operator $\cT$ is a contraction when restricted to the subset $\cL_\gamma$ of $Q$-functions such that "$Q_r$ is Lipschitz with respect to $Q_c$":
\begin{equation}
    \cL_\gamma = \left\{\begin{array}{cc}
         Q\in(\Real^2)^{\ocS\ocA}\text{ s.t. }\exists L<\frac{1}{\gamma}-1: \forall \os\in\ocS,\oa_1,\oa_2\in\ocA,   \\
         |Q_r(\os,\oa_1) - Q_r(\os,\oa_2)| \leq L|Q_c(\os,\oa_1) - Q_c(\os,\oa_2)|
    \end{array}\right\}
\end{equation}
\end{theorem}

Thus, we expect that \Cref{algo:bvi} is likely to converge when $Q^*$ is smooth, but could diverge if the slope of $Q^*$ is too high.  $L^2$-regularisation can be used to encourage smoothness and mitigate risk of divergence.

@@

We first bring to light the role of risk-sensitive exploration in the \text{corridors} environment: \Cref{fig:exploration} shows the set of trajectories collected by each exploration strategy.
and the resulting performance of a budgeted policy trained on each batch. The trajectories (orange) in the risk-neutral batch are concentrated along the risky corridor (right) and ignore the safe corridor (left), which results in bad performances in the low-risk regime. Conversely, trajectories in the risk-sensitive batch (blue) are well distributed among both corridors and the corresponding budgeted policy achieves good performance across the whole spectrum of risk budgets.

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.49\textwidth]{source/img/corridors_densities.pdf}
    %\includegraphics[width=0.23\textwidth]{source/img/risk-sensitive.png}
    \includegraphics[page=1, width=0.49\textwidth]{source/img/corridors}
    \caption{Density of explored states (left) and corresponding policy performances (right) of two exploration strategies in the {corridors} environment. }
    \label{fig:exploration}
\end{figure}

@@

In a second experiment displayed in \Cref{fig:results}, we compare the performance of \FTQl to that of \BFTQ in the dialogue and autonomous driving tasks. For each algorithm, we plot the reward-cost trade-off curve. In both cases, \BFTQ performs almost as well as \FTQl despite only requiring a single model. All budgets are well-respected on \text{slot-filling}, but on \text{highway-env} we can observe an underestimation of $Q_c$, since e.g. $\expectedvalue[G_c|\beta=0] \simeq 0.1 $. This underestimation can be a consequence of two approximations: the use of the sampling operator $\hat{\cT}$ instead of the true population operator $\cT$, and the use of the neural network function approximation $Q_\theta$ instead of $Q$. 
% This is probably due to \text{highway-env} being actually infeasible for $\beta=0$, since we noticed that about $8\%$ of bad initialisations occur with the front vehicle driving slower than the minimum velocity of the controlled vehicle, leading to either a front collision or a forced change to the opposite lane, both yielding costs.
Still, \BFTQ provides a better control on the expected cost of the policy, than \FTQl. In addition, \BFTQ behaves more consistently than \FTQl overall, as shown by its lower extra-seed variance.

Additional material such as videos of policy executions is provided in \Cref{sec:exp-supp}.

### 2019-10-26 paper/poster/poster.tex
Updates from Overleaf

\end{minipage}
\hfill%
\begin{minipage}{0.25\textwidth}
\begin{algorithm}[H]
\DontPrintSemicolon
\KwData{$P, R_r, R_c$}
\KwResult{$Q^*$}
$Q_{0} \leftarrow 0$\;
\Repeat{convergence}{
    $Q_{k+1} \leftarrow \cT Q_k$\;
}
\caption{Budgeted Value Iteration}
\label{algo:bvi}
\end{algorithm}
\begin{center}
\includegraphics[width=0.9\textwidth]{source/img/concavity_example.pdf}    
\end{center}
\end{minipage}

@@

\bookboxx{
\begin{remark*}[Contractivity on smooth $Q$-functions]
\label{rmk:contractivity-smooth}
We conjecture that \hlg{$\cT$ is a contraction when restricted} to the subset $\cL_\gamma$ of $Q$-functions such that "$Q_r$ is \hlg{$L$-Lipschitz} with respect to $Q_c$", with $L<\frac{1}{\gamma}-1$.
\end{remark*}
}
}

@@

\includegraphics[width=0.3\textwidth]{source/img/test.pdf}
\includegraphics[page=1, width=0.49\textwidth]{source/img/corridors}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\centering
Pareto frontier

>>>>

\end{minipage}
\hfill%
\begin{minipage}{0.25\textwidth}
\begin{algorithm}[H]
\DontPrintSemicolon
\KwData{$P, R_r, R_c$}
\KwResult{$Q^*$}
$Q_{0} \leftarrow 0$\;
\Repeat{convergence}{
    $Q_{k+1} \leftarrow \cT Q_k$\;
}
\caption{Budgeted Value Iteration}
\label{algo:bvi}
\end{algorithm}
\begin{center}
\includegraphics[width=0.9\textwidth]{../source/img/concavity_example.pdf}    
\end{center}
\end{minipage}

@@

\bookboxx{
\begin{theorem*}[Contractivity on smooth $Q$-functions]
\label{rmk:contractivity-smooth}
\hlg{$\cT$ is a contraction when restricted} to the subset $\cL_\gamma$ of $Q$-functions such that "$Q_r$ is \hlg{$L$-Lipschitz} with respect to $Q_c$", with $L<\frac{1}{\gamma}-1$.
\end{theorem*}
}
}

@@

\includegraphics[width=0.49\textwidth]{../source/img/corridors_densities.pdf}
\includegraphics[page=1, width=0.49\textwidth]{../source/img/corridors}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\centering
Pareto frontier

### 2019-10-26 paper/supplementary.tex
Updates from Overleaf

@@

%\begin{minted}{sh}
\begin{lstlisting}[language=bash,caption={bash version}]
# Install highway-env
pip3 install --user git+https://github.com/eleurent/rl-agents
# clone the phd_code repository
git clone https://github.com/ncarrara/phd_code
# Change python path to the path of this repository
export PYTHONPATH="phd_code/ncarrara"
# Navigate to budgeted-rl folder
cd phd_code/ncarrara/budgeted-rl/
# Run main script using any config file
# Choose the range of seeds you want to test on
python3 main/egreedy/main-egreedy.py config/slot-filling.json 0 6
python3 main/egreedy/main-egreedy.py config/corridors.json 0 4
python3 main/egreedy/main-egreedy.py config/highway-easy.json 0 10
\end{lstlisting}
%\end{minted}

@@

    \begin{itemize}\item \textbf{yes}, see \Cref{sec:intro}, \Cref{sec:bdp}, \Cref{sec:brl},
    \Cref{sec:exploration} and \Cref{sec:bftq-full}.
    \end{itemize}
    \item An analysis of the complexity (time, space, sample size) of any algorithm:
    \begin{itemize}\item \textbf{yes}, see \Cref{subsec:parallel-computing}.
    \end{itemize}
    \item A link to a downloadable source code, with specification of all dependencies, including external libraries:
    \begin{itemize}\item \textbf{yes}, see \Cref{subsubsec:instruction-reproducibility} and the folder \texttt{code} in the supplementary material zip file.
    \end{itemize}
\end{itemize}

@@

        \begin{itemize}
            \item  \textbf{yes}, two environments   are shipped with the supplementary material (in the \texttt{code} folder) and the third one is fetch from a public repository, see \Cref{subsubsec:instruction-reproducibility} for details.
        \end{itemize}
    \item An explanation of any data that were excluded, description of any pre-processing step: 
        \begin{itemize}
            \item it's \textbf{not applicable} as data comes from simulated environments, so pre-processing steps are not needed.
        \end{itemize}

@@

        \begin{itemize}
            \item  \textbf{yes}, we even observe less variability with our novel approach, see \Cref{subsec:results}.
        \end{itemize}
    \item A description of the computing infrastructure used:     \begin{itemize}
            \item The Graphic Processing Unit used for experiments is an NVIDIA \texttt{GeForce GTX 1080 Ti} and the Computational Processing Unit is an Intel \texttt{Xeon E7}.
        \end{itemize}
\end{itemize}

>>>>

\begin{remark}
Because of the budget dynamics $\beta' = \beta_a$ in \eqref{eq:dynamics}, we can see in \eqref{eq:bellman_expectation} that for all $\os=(s, \beta)$ and $\oa=(a, \beta_a)$, $Q^\pi(\os, \oa)$ only depends on $s, a, \beta_a$ and not on $\beta$. We will slightly abuse notations and sometimes denote in the following $Q^\pi(s, a, \beta_a) \eqdef Q^\pi(\os, \oa)$.
\end{remark}

@@

%\begin{minted}{sh}
\begin{lstlisting}[language=bash,caption={bash version}]
# Install highway-env
pip3 install --user git+https://github.com/eleurent/rl-agents
# clone the phd_code repository
git clone https://github.com/ncarrara/budgeted-rl
# Change python path to the path of this repository
export PYTHONPATH="budgeted-rl/ncarrara"
# Navigate to budgeted-rl folder
cd budgeted-rl/ncarrara/budgeted-rl
# Run main script using any config file
# Choose the range of seeds you want to test on
python3 main/egreedy/main-egreedy.py config/slot-filling.json 0 6
python3 main/egreedy/main-egreedy.py config/corridors.json 0 4
python3 main/egreedy/main-egreedy.py config/highway-two-way.json 0 10
\end{lstlisting}
%\end{minted}

@@

    \begin{itemize}\item \textbf{yes}, see \Cref{sec:intro}, \Cref{sec:bdp}, \Cref{sec:brl},
    \Cref{sec:exploration} and \Cref{sec:bftq-full}.
    \end{itemize}
    \item An analysis of the complexity (time, space, sample size) of any algorithm:
    \begin{itemize}\item \textbf{yes}, see \Cref{subsec:parallel-computing}.
    \end{itemize}
    \item A link to a downloadable source code, with specification of all dependencies, including external libraries:
    \begin{itemize}\item \textbf{yes}, see \Cref{subsubsec:instruction-reproducibility}. All resources are available at \href{https://budgeted-rl.github.io}{this https URL}.
    \end{itemize}
\end{itemize}

@@

        \begin{itemize}
            \item  \textbf{yes}, all environments  are fetched from public repositories, see \Cref{subsubsec:instruction-reproducibility} for details.
        \end{itemize}
    \item An explanation of any data that were excluded, description of any pre-processing step: 
        \begin{itemize}
            \item it's \textbf{not applicable} as data comes from simulated environments, so pre-processing steps are not needed.
        \end{itemize}

@@

        \begin{itemize}
            \item  \textbf{yes}, we even observe less variability with our novel approach, see \Cref{subsec:results}.
        \end{itemize}
    \item A description of the computing infrastructure used:     \begin{itemize}
            \item The Graphic Processing Unit used for experiments is an NVIDIA \texttt{GeForce GTX 1080 Ti} and the Central Processing Unit is an Intel \texttt{Xeon E7}.
        \end{itemize}
\end{itemize}
